Exercise 5. Dimensionality reduction techniques
---
```{r}
#Read the data and show summaries of the variables.
human<-read.table(file = "human.txt", sep="\t", header=TRUE)
str(human)
summary(human)
```

There are 8 variables and 155 observations.


```{r}
#Access GGally.
library(GGally)

#Visualize the variables.
ggpairs(human)
```

Many of the variables are skewed to either side and/or have long tails. Education expectancy looks the most normally distributed.
Mean education expectancy for example is 13.2 years, mean life expectancy 71.7 years and mean adolescent birth rate 47.2. 
There is great variation within some variables: For the Gross national income per capita, the minimum value is 581, median 12040 and maximum 123124 and for the Maternal mortality ratio, the minimum value is 1, median 49 and maximum 1100. Then, in the Education expectancy variation is relatively small as the minimum value is 5.4, median 13.5 and maximum value 20.2.

There are several variables that are highly correlated with each other, for example:
Adolescent birth rate and Maternal mortality,
Life expectancy and Education expectancy,
Life expectancy and Adolescent birth rate,
Education expectancy and Maternal mortality.

Percentage of female representatives in parliament and Adolescent birth rate, for example, are not correlated.


### Principal component analysis for non-standardized data

```{r}
#Perform principal component analysis on the non-standardized data.
pca_human <- prcomp(human)
```


```{r}
#Create and print out a summary of pca_human.
s <- summary(pca_human)
s
```


```{r}
# Rounded percetanges of variance captured by each PC.
pca_pr <- round(1*s$importance[2,]*100, digits = 1) 
```


```{r}
#Print out the percentages of variance.
pca_pr
```


```{r}
#Create object pc_lab to be used as axis labels.
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)") 
```


```{r}
#Draw a biplot.
biplot(pca_human, cex = c(0.8, 0.1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])
```


### Principal component analysis for standardized data

```{r}
#Standardize the variables.
human_std <- scale(human)
```


```{r}
#Print out summaries of the standardized variables.
summary(human_std)
```


```{r}
#Perform principal component analysis (with the SVD method).
pca_human <- prcomp(human_std)
```


```{r}
#Create and print out a summary of pca_human.
s <- summary(pca_human)
s
```


```{r}
#Rounded percetanges of variance captured by each PC.
pca_pr <- round(1*s$importance[2,]*100, digits = 1)
```

 
```{r}
#Print out the percentages of variance.
pca_pr
```


```{r}
#Create object pc_lab to be used as axis labels.
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)") 
```


```{r}
#Draw a biplot.
biplot(pca_human, choices = 1:2, cex = c(0.6, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2], main = "Global health and equality related indicators and their connections") 

```


In the non-standardized data, the first principal component PC1 explains 100% of the variance and the second principal component PC2 explains 0% of the variance. In the standardized data, PC1 explains 53.6% of the variance and PC2 explains 16.2% of the variance. Together they explain 69.8% of the variance of the whole dataset.
In PCA, the principal components are constructed so, that the first principal component accounts for the largest possible variance in the data set and the second one the second largest possible variance.

In the non-standardized data, there were variables with high range (great variance), which dominated the ones with smaller range, which lead to biased results. Standardization standardizes the variables so that every one of them contributes equally to the analysis, which prevents the problem of variables with larger range dominating and causing biased results.

In the non-standardized data, for example Gross national income GNI had such a big variance that it caused biased results. After standardization and removing the domination problem, the effects of the variables with smaller variation, for example education expectancy, can be seen.



### Personal interpretations of the first two principal component dimensions

Iceland, Sweden, Norway, Finland and Denmark are closely grouped together and are therefore similar to each other.
They all have quite high Percentage of female representatives in parliament, Expected years of education, Gross national income per capita, Population with secondary Education female to male ratio and life expectancy at birth. 

The arrows visualize the connections between the original variables and the principal components.
Expected years of education, Gross national income per capita, Population with secondary education female to male ratio and Life expectancy at birth have a small angle between them, so we can assume correlation between them. Maternal mortality ratio and adolescent birth rate also seem to be correlated.
The angle of the arrows for Percentage of female representatives in parliament and Adolescent birth rate for example is quite large and therefore we can assume that the is no correlation between those variables.

Expected years of education, Gross national income per capita, Population with secondary education female to male ratio, Life expectancy at birth, Maternal mortality ratio and Adolescent birth rate all have a small angle with the principal component 1 and can be assumed to correlate with each other.
Percentage of female representatives in parliament and Labour force participation rate female/male ratio on the other hand, can be assumed to correlate with principal component 2.

Sierra Leone, Liberia, Congo and Chad for example seem to have both high Maternal mortality ratio and Adolescent birth rate and do not have high Expected years of education, Gross national income per capita or life expectancy at birth. Mosambique has both quite high Maternal mortality ratio and Adolescent birth rate but women also have a high Labour force participation rate.



### Tea dataset and multiple correspondence analysis


```{r}
#Load library and tea dataset.
library(FactoMineR)
data(tea)
```


```{r}
#Explore tea dataset.
dim(tea)
str(tea)
summary(tea)
```

There are 36 variables and 300 observations. All the variables are factorial, except age, which is integral.


```{r}
#Access library dplyr.
library(dplyr)
library(tidyr)

#Column names to keep in the dataset.
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

#Select the 'keep_columns' to create a new dataset.
tea_time <- dplyr::select(tea, one_of(keep_columns))

#Look at the summaries and structure of the data.
str(tea_time)
summary(tea)
```

There are now 6 variables and 300 observations. 

```{r}
#Multiple correspondence analysis.
mca <- MCA(tea_time, graph = FALSE)

#Summary of the model.
summary(mca)
```

```{r}
#Visualize the dataset.
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

#Visualize Multiple Correspondence Analysis.
plot(mca, invisible=c("ind"), habillage="quali")
```

People who drink unpackaged tea buy it from tea shops and people who use tea bags for theit tea buy it from the chain stores. People who drink both unpackaged tea and use tea bags also buy their tea from both types of shops.




