Exercise 4 Clustering and classification.
---

```{r}
#Access the MASS package.
library(MASS)
```

```{r}
#Load the data.
data("Boston")
```


```{r}
#Explore the dataset.
str(Boston)
summary(Boston)
```


```{r}
#Plot matrix of the variables.
pairs(Boston)
```

There are 506 observations (rows) and 14 variables (columns) in the dataset. The variables are for example crim (per capita crime rate by town), zn (proportion of residential land zoned for lots over 25,000 sq.ft), indus (proportion of non-retail business acres per town), chas (Charles River dummy variable (= 1 if tract bounds river; 0 otherwise)), nox (nitrogen oxides concentration (parts per 10 million)), lstat (lower status of the population (percent)) and medv (median value of owner-occupied homes in \$1000s).

For more details of the dataset, please visit  https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html

```{r}
library(dplyr)
library(corrplot)
```


```{r}
#Calculate the correlation matrix and round it.
cor_matrix<-cor(Boston) %>% round(digits = 2)
```


```{r}
#Print the correlation matrix.
cor_matrix
```


```{r}
#Visualize the correlation matrix.
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6)
```

Positive correlations are displayed in blue and negative correlations in red color. Color intensity and the size of the circle are proportional to the correlation coefficients.
There is a high negative correlation between indus (proportion of non-retail business acres per town) and dis (weighted mean of distances to five Boston employment centres), nox (nitrogen oxides concentration (parts per 10 million)) and dis (weighted mean of distances to five Boston employment centres), age (proportion of owner-occupied units built prior to 1940) and dis (weighted mean of distances to five Boston employment centres) and istat (lower status of the population (percent)) and medv (median value of owner-occupied homes in \$1000s).
There is a high positive correlation between rad (index of accessibility to radial highways) and tax (full-value property-tax rate per \$10,000). 


Standardize the dataset, create a categorical variable of the crime rate, divide the dataset to train and test sets.


```{r}
#Center and standardize variables.
boston_scaled <- scale(Boston)
```


```{r}
#Summaries of the scaled variables.
summary(boston_scaled)
```


```{r}
#Class of the boston_scaled object.
class(boston_scaled)
```


```{r}
#Change the object to data frame.
boston_scaled <- as.data.frame(boston_scaled)
```

After scaling the mean is 0 for all the variables which means that all variables are normally distributed. 

```{r}
#Summary of the scaled crime rate.
summary(boston_scaled$crim)
```

```{r}
#Create a quantile vector of crim and print it.
bins <- quantile(boston_scaled$crim)
bins
```


```{r}
#Create a categorical variable 'crime'.
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels = c("low", "med_low", "med_high", "high"))
```

```{r}
#Look at the table of the new factor crime.
table(crime)
```


```{r}
#Remove original crim from the dataset.
boston_scaled <- dplyr::select(boston_scaled, -crim)
```


```{r}
#Add the new categorical value to scaled data.
boston_scaled <- data.frame(boston_scaled, crime)
```


Divide the dataset to train and test sets, so that 80% of the data belongs to the train set.

```{r}
#Number of rows in the Boston dataset. 
n <- nrow(boston_scaled)
```


```{r}
#Choose randomly 80% of the rows.
ind <- sample(n,  size = n * 0.8)
```


```{r}
#Create train set.
train <- boston_scaled[ind,]
```


```{r}
#Create test set.
test <- boston_scaled[-ind,]
```


```{r}
#Save the correct classes from test data.
correct_classes <- test$crime
```


```{r}
#Remove the crime variable from test data.
test <- dplyr::select(test, -crime)
```

Fit the linear discriminant analysis on the train set.

```{r}
#Linear discriminant analysis.
lda.fit <- lda(crime ~ ., data = train)
```


```{r}
#Print the lda.fit object.
lda.fit
```


```{r}
#The function for lda biplot arrows.
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}
```


```{r}
#Target classes as numeric.
classes <- as.numeric(train$crime)
```


```{r}
#Plot the lda results.
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)
```

The best linear separator is variable index of accessibility to radial highways (rad), which has the longest arrow in the picture. The second best separator is difficult to make out, but it looks like it would be nitrogen oxides concentration (parts per 10 million) (nox).


Predicting with the model.

```{r}
#Predict classes with test data.
lda.pred <- predict(lda.fit, newdata = test)

#Cross tabulate the results.
table(correct = correct_classes, predicted = lda.pred$class)
```


Note that the numbers change every time you run the model.
The model predicted the high crime rates well (32/32 were classified correctly). Other categories the model did not predict as well:
For medium high crime rates, 17/29 were classified correctly, for medium low crime rates 17/25 were classified correctly, and for low crime rates 9/16 were classified correctly.


```{r}
total <- c(9+5+2+3+17+5+10+17+2+32)
total

correct <- c(9+17+17+32)
correct
```

Out of a total of 102 observations, 75 observations were classified correctly. 


```{r}
ratio <- c(correct/total)
ratio
```

Accuracy of the model was 74%, which is not that bad but could be better.


Reload Boston dataset.

```{r}
library(MASS)
data("Boston")
```


```{r}
#Center and standardize variables.
boston_scaled <- scale(Boston)
```


```{r}
#Change the object to data frame from matrix type.
boston_scaled <- as.data.frame(boston_scaled)
```


```{r}
#Calculate the Euclidean distances between observations.
dist_eu <- dist(boston_scaled)
```


```{r}
#Look at the summary of the distances.
summary(dist_eu)
```

Run k-means algorithm on the dataset. 
```{r}
#K-means clustering.
km <-kmeans(boston_scaled, centers = 3)
```


```{r}
#Plot the Boston dataset with clusters.
pairs(boston_scaled, col = km$cluster)
```


```{r}
#Investigate the optimal number of clusters and run the algorithm again.
set.seed(123)
```


```{r}
#Determine the number of clusters.
k_max <- 10
```


```{r}
#Calculate the total within sum of squares.
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled, k)$tot.withinss})
```


```{r}
#Visualize the results with qplot. Visualize (with qplot) the total WCSS when the number of cluster goes from 1 to 10.
library(ggplot2)
qplot(x = 1:k_max, y = twcss, geom = 'line')
```

2 clusters seems optimal as the bend (knee) is at 2.


```{r}
#Run kmeans() again with two clusters.
km <-kmeans(boston_scaled, centers = 2)
```

```{r}
#Plot the Boston dataset with clusters.
pairs(boston_scaled, col = km$cluster)
```

Like observed before, the optimal number of clusters seems to be two.


Super bonus.

```{r}
model_predictors <- dplyr::select(train, -crime)
```

```{r}
#Check the dimensions.
dim(model_predictors)
dim(lda.fit$scaling)
```


```{r}
#Matrix multiplication.
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)
```

```{r}
#Matrix multiplication.
library(plotly)
```

```{r}
#Create a 3D plot of the columns of the matrix product by typing the code below.
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers')
```

```{r}
#Add argument color as an argument in the plot_ly() function.
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = train$crime)
```


Draw another 3D plot where the color is defined by the clusters of the k-means.  

```{r}
#Make a k-means with 4 clusters to compare the methods.
km3D <-kmeans(boston_scaled, centers = 4)
```

```{r}
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers', color = km3D$cluster[ind])
```

Medium high crime rates seems to bee better defined than cluster 1. Cluster 2 is a bit better defined (not so much intermingling) than low crime rate in the first picture.  Cluster 3 is quite similar with high crime rates, even though cluster 3 is a bit more better defined. Cluster 4 is more defined than medium low crime rates in the first picture. 

