Exercise 6. Analysis of longitudinal data
---

#### Implement the analyses of Chapter 8 of MABS using the RATS data. 

```{r}
#Read the data and explore dimensions.
RATSL<-read.table(file = "RATSL.txt", sep="\t", header=TRUE)

dim(RATSL)
str(RATSL)
summary(RATSL)
```

```{r}
#Change the categorical variables to factors.
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

dim(RATSL)
str(RATSL)
summary(RATSL)
```

There are 176 observations and 5 variables. Data was converted to long form in the data wrangling exercise.
Rat's weights have been measured several times on various weekdays.

```{r}
#Plot the rat weights.
library("tidyverse")

ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID))+
  geom_line() + scale_linetype_manual(values = rep(1:10, times=4))+
  facet_grid(. ~ Group, labeller = label_both)+
  theme_bw() + theme(legend.position = "none")+
  theme(panel.grid.minor.y = element_blank()) +
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

The plot shows 3 different groups and the rat's weight change over time. In group 1, the weight of the rats is lower than in the other groups.
The weights of the rats is highest in the group 3. All groups seem to have an outlier, in group 2 the "biggest" one.
In all groups the weight of the rats increases during the experiment. 

```{r}
#Standardize the variable weight.
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = Weight) %>%
  ungroup()
```


```{r}
#Glimpse the data and check the structure.
glimpse(RATSL)
str(RATSL)
summary(RATSL)
```


```{r}
#Plot the standardized rat weights.
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() + scale_linetype_manual(values = rep(1:10, times=4))+
  facet_grid(. ~ Group, labeller = label_both)+
  theme_bw() + theme(legend.position = "none")+
  theme(panel.grid.minor.y = element_blank())+
  scale_y_continuous(name = "Standardized weight")
```

The plot shows 3 different groups and the rat's weight change over time after standardization.


### Plot showing average (mean) to make graph more readable.

```{r}
#Number of weeks, baseline (week 0) included.
n <- RATSL$Time %>% unique() %>% length()
#Summary data with mean and standard error of weight by group and time.
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), sd = sd(Weight)) %>%
  ungroup()
```


```{r}
# Glimpse the data
glimpse(RATSS)
```



```{r}
#Plot the mean profiles.
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  #geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

The plot shows 3 groups and the mean weight change over time.
There is no overlap in the mean profiles of the three groups, which suggests that there is a difference between the three groups with respect to the mean weight values.


#### Applying the Summary Measure Approach

```{r}
#Create a summary data by Group and ID with mean as the summary variable (ignoring baseline time 1).
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
```


```{r}
#Glimpse the data.
glimpse(RATSL8S)
```


```{r}
#Draw a boxplot of the mean versus treatment.
library(dplyr)
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weeks 1-8")
```


We can see one outlier in each group. In group 2, the outlier is furthest distance from the other values. In groups 1 and 3 the sistance of the outlier from other values is not as great.

#### Draw the boxplot without the outliers

#Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data.

```{r}
#Draw a boxplot of the mean versus treatment.
#Find the numerical values for the outliers.
RATSL8S$mean
```


```{r}
#Filter the outliers.
RATSL8S1 <- filter(RATSL8S, mean != 238.9, mean != 594.0, mean != 495.2)

RATSL8S1$mean
```


```{r}
#Create a boxplot.
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), weeks 1-8")
```


After the outliers are removed, the variation inside the groups decreases notably, which can be seen from the picture.
As there are three groups instead of two, t-test cannot be used.

```{r}
#Add the baseline from the original data as a new variable to the summary data.
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = filter(RATSL, Time==1)$Weight)
```


```{r}
#Fit the linear model with the mean as the response. 
fit <- lm (mean ~ Group, data = RATSL8S2)
summary(fit)
```

```{r}
#Compute the analysis of variance table for the fitted model with anova().
anova(fit)
```

There is a statistically significant difference (p<0.001) between groups 2 and group 3 compared with group 1. 




